Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-02-01 paper/paperSource.tex
paper changes

%\smartqed  % flush right qed marks, e.g. at end of proof

%----------------------------------------------------------

@@

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/speedup838K-pomalo-za-4-eps-converted-to.pdf} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainPaper}}
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

\section{Related Work}\label{sc:relatedwork}
The idea of data streaming has been broadly used in many fields. The
main usage however is aimed at processing live data generated
online. There is an ocean of references for data stream management,
but we believe the presentation in~\cite{DataStreamManagement}
encapsulates the majority of the ideas, problems and solutions. In
the past decade, however, the database community started to follow
the data stream idea to process stored data. Processes can take
advantage of the streaming data at any time the data is being pushed
through the system. This gives rise to push-based type design for
data management systems. The idea of such push-based design was
previously considered in projects such as DataPath~\cite{DataPath},
Volcano~\cite{Volcano} and QPipe~\cite{Qpipe} among others. These
works show ideas in which the data-driven dataflow is compared to
the demand-driven dataflow showing the need for the later. They also
talk about maximizing the data and work sharing among queries at
runtime. Essentially, we incorporate such ideas in our design of the
push-based system.

On the other side, the scientific community has steadily progressed
from processing massive data files towards employing database
systems for the storage, acquisition, and analysis of large-scale
scientific data~\cite{SDSS_SIGMOD02,QBISM_ICDE94}. The widely used
and popular relational database systems are conventionally designed
and optimized to better manage the data produced by the business
type applications. But such conventional database systems (DBMS) are
not well equipped to deal with the type and quantity of scientific
data, such as the data produced by the molecular simulations. In the
recent past, the DBMS community has made some attempts into the
design and construction of database systems optimized for handling
scientific data. Such examples include the BDBMS
project~\cite{BDBMS_CIDR07} that deals with annotation and
provenance of the sequence data in biosciences. and the PeriScope
project~\cite{PeriScope} is designed to efficiently handle
declarative queries against bio sequences. On top of the
aforementioned examples, there are also ideas for new DBMS
frameworks aimed at the management of scientific
data~\cite{NewArch-VLDB07,ROI-CIDR07,SciDB-SIGMOD10}. One of those
systems is the SciDB~\cite{SciDB-SIGMOD10,SciDB} and it is closest
to the idea presented in this paper. SciDB is data management and
analytics system that is primarily used is in application domains
involving very big scale array data. This system, like the one
presented in this paper is designed around a multi-dimensional array
data-model and it uses arrays to store the data. SciDB stores
petabytes of data on a number of machines and runs its queries on
those machines. It is made for high performance, high-availability,
fault tolerance, and scalability. However, to the best of our
knowledge, it too follows the pull based design where its queries
demand the data they need. As seen earlier in this paper, this type
of design can impose I/O overhead and decrease the data throughput
when doing the analysis. Aside the mentioned issue, the design and
build of such DBMS optimized for scientific data management come
with a additional challenges. Such challenges as well as their
probable resolution are outlined in~\cite{Sci_SIGMODR05}. Recently,
there have been some efforts aimed at designing and building MS data
managements systems on top of relational databases. Such efforts are
presented through projects like BioSimGrid~\cite{BioSimGrid} and
SimDB~\cite{SimDB} that were developed especially for molecular
simulations. However, to the best of our knowledge, such systems
still lack the efficiency needed for MS data management as well as
efficient query processing strategies.

Generally, the data produced in the process of molecular simulation
is being stored in large, plain files with no structure whatsoever.
Queries, which are implemented in a stand alone programs within
simulation/analysis systems, are executed onto such files producing
the quantities that scientist use to analyze the molecular system.
Such simulation and/or analysis packages include: Gromacs, VMD,
MDAnalysis, Wordom, MD-TRACKS, SimulaidOne, Charmm among the others.
But to the best of our knowledge, all of these systems work on a
similar basis: they take a user defined query and execute it onto
the MS simulated data. In order for the query to be executed, the
data has to be loaded into the main memory. Then the result is
either produced onto the display or written to a file. When the next
user query comes, the system again loads significant part of the
dataset into the main memory and executes the query. We believe that
there is a room for improvement of such systems, given the fact that
many of the user defined queries executed during system's analysis
are fairly static. In other words, there is a number of queries that
a user would always want to execute on a given simulation data.
Furthermore, the selections of MS data onto which such queries might
be executed, are also fairly constant (i.e., oftentimes the user
selects the same group of atoms (e.g., all hydrogen atoms) to
calculate given quantity, like center of mass for instance). So, by
precoding many such queries and running them automatically once the
system has loaded the data into memory, we believe we can save a lot
of time that otherwise would have been spent in loading the same
data into main memory anytime a query is executed. On top of the
automated query execution, our system can take user's query as input
as well. With this, we believe our system is an improvement over the
MS analysis systems that are used today.

>>>>

%\smartqed  % flush right qed marks, e.g. at end of proof
%----------------------------------------------------------

@@

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/speedup838K-pomalo-za-4.eps} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainPaper}}
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

\vspace{3mm}
\section{Related Work}\label{sc:relatedwork}
\vspace{3mm}

Push-based system for MS data analysis is primarily proposing an idea of pushing chunks of data through a network of queries. Essentially, this has also been proposed, as mentioned already, by other frameworks \cite{DataPath,Volcano,Qpipe}. In other words, the major point is to use common data loaded into memory for execution of concurrent queries, which is supposed to eliminate I/O overhead predominantly. The architecture and explanation of DSMS (Data Stream Management System) versus DBMS is well explained in one of the lectures of Morgan and Claypool \cite{DataStreamManagement}. It is very important to capture the differences and specialties of DSMS in network architecture, stream models and windows, scheduling, load balancing, approximation, data expiration, etc.

Since dealing with data is very delicate in this kind of engines, there are couple of features and requirements that this architecture is applicable to, in order to avoid inappropriate understanding or application of this approach in systems that were covered in the lectures\cite{DataStreamManagement}. The data is volatile and not persistent anymore because of obvious sizes of data.  Since we are pulling the data as little times as possible taking into consideration that MS data is only appended and not modified, the access is sequential rather than random. There are concepts of framing and windowing, and memory is limited in streaming engine, while in DBMS secondary storage is considered to be unlimited. Obviously, because of all of these features or requirements to data, the streaming engine frameworks have some limitations like going back into the history of data, since it is volatile. In fact, this has also been tried by Borealis Streaming Engine \cite{borealis}. On top of the fact that it is a streaming engine, as it was built based on Medusa\cite{medusa} and Aurora\cite{aurora}, it proposes distributed processing at this point having some features to handle errors and look back in the history. In our case, it is primarily a single data source, since it is based on the simulation, though in the future the simulation software systems might introduce distributed computing, too. It is also important to mention such projects as PeriScope\cite{PeriScope}, SciDB\cite{SciDB}, and other\cite{SDSS_SIGMOD02,QBISM_ICDE94}, since they target scientific data, which is a lot more different, than usual industry customer oriented data.

This version of Push-based System of MS data analysis is primarily about taking an advantage of GPU processing of the data. Hadoop is a well-known and well-used framework by many companies nowadays\cite{hadoop}. Its purpose is an ability to store and serve large scale data across clusters in a timely fashion. Since usage of GPU of massive data (not only related to graphics processing) is a relatively new concept, there is also an improved version of Hadoop MapReduce Framework with GPU\cite{hadoopgpu}. Having an ability to scale it up in a distributed computing environment is probably one of the best approaches, but again it might be a bigger overhead in terms of costs, since keeping all nodes up and still being able to solve specific problem sets with hardware stack of a lower capacity raise a will to explore more. Having a specific software for simulation like Gromacs, apparently there is also a possibility of taking an advantage of GPUs. For example, Gromacs allows to optimize simulation and query tools by adding configurations based on GPU device located on the processing computer. Unfortunately, even with this feature, Gromacs doesn't follow push-based approach which means that improvement with GPU with pull data per query makes it negligible.

Gromacs generate output files with physical properties description for each atom in each frame. Modern CPUs do great job in terms of caching and computing, but obviously in this particular problem massive monolithic computation is needed, thus, GPU devices are perfect candidates for such a problem. To be more exact, for example, a single kernel in a GPU device could be dedicated to a single atom (this relationship might change depending on the query, but this is to understand the scale and relationship of multiprocessors), and since we have scientific data of MS, the data is appended, which means that we will only move forward and not take an advantage of caching on chunks high level (though we might take an advantage of GPU caching features particularly in processing phase) having multiple workloads per chunk. Just to make it more clear, workloads in this particular case represent transactions based on number of clients. For example, if we were to serve it all in a streaming fashion, 100 clients might demand concurrent queries with different slight selection properties and expect their results, each workload might be dedicated to each client if not aggregated. Thus, the performance of GPU is certainly increasing based on workload too, which might benefit considerably.
